simodels: A package for Spatial Interaction Modelling

### 2024-09-02 paper/paper.qmd
Updates after conversation with @adamdennett

---
title: "simodels: A package for Spatial Interaction Modelling"
# Note: you need to have the relevant quarto extension installed:
# quarto add quarto-journals/jss
# After that put the following in the yaml:
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
# author:
#   - name: Achim Zeileis
#     affiliations:
#       - name: Universität Innsbruck
#         department: Department of Statistics
#         address: Universitätsstr. 15
#         city: Innsbruck
#         country: Austria
#         postal-code: 6020
#       - Journal of Statistical Software
#     orcid: 0000-0003-0918-3766
#     email: Achim.Zeileis@R-project.org
#     url: https://www.zeileis.org/
#   - name: Second  Author
#     affiliations:
#       - Plus Affiliation
# abstract: |
#   This short article illustrates how to write a manuscript for the
#   _Journal of Statistical Software_ (JSS) using its {{< latex >}} style files.
#   Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
#   it is recommended to keep the {{< latex >}} code as simple as possible,
#   i.e., avoid inclusion of packages/commands that are not necessary.
#   For outlining the typical structure of a JSS article some brief text snippets are employed that have been inspired by @ZeileisKleiberJackman2008, discussing count data regression in [R]{.proglang}. Editorial comments and instructions are marked by vertical bars.
abstract: |
  This article introduces the [simodels]{.pkg} package for spatial interaction modelling in R. The package provides functions for estimating spatial interaction models, including unconstrained gravity models and models that are production-, attraction-, or doubly-constrained.

keywords: [JSS, style guide, comma-separated, not capitalized, R]
keywords-formatted: [JSS, style guide, comma-separated, not capitalized, "[R]{.proglang}"]

bibliography: bibliography.bib
execute: 
  messages: false
  warning: false 
  echo: false
---

-->

@@

## Basic implementation of SIMs {#sec-base}

@@

## OD data preparation {#sec-od}

## Interaction calculation {#sec-interaction}

## Interaction modelling {#sec-models}

@@

### Commuter flows in Leeds, UK

@@

```{r}
od_modelled_max_5km = si_to_od(zones, destinations)
od_modelled = si_to_od(zones, destinations, max_dist = 5000)
dim(od_modelled)
names(od_modelled)
```

>>>>

---
title: "simodels: A package for Spatial Interaction Modelling"
bibliography: bibliography.bib
execute: 
  messages: false
  warning: false 
  echo: false
---

```{r}
#| eval: false
#| echo: false
# Compile to docx and jss-pdf formats:
quarto::quarto_render("paper/paper.qmd", output_format = "docx")
quarto::quarto_render("paper/paper.qmd", output_format = "jss-pdf")
system("gh release upload v0.1.0 paper/paper.docx --clobber")
system("gh release upload v0.1.0 paper/paper.pdf --clobber")
```

<!-- 
# Note: you need to have the relevant quarto extension installed for the JSS version:
# quarto add quarto-journals/jss
# After that put the following in the yaml:
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
# author:
#   - name: Achim Zeileis
#     affiliations:
#       - name: Universität Innsbruck
#         department: Department of Statistics
#         address: Universitätsstr. 15
#         city: Innsbruck
#         country: Austria
#         postal-code: 6020
#       - Journal of Statistical Software
#     orcid: 0000-0003-0918-3766
#     email: Achim.Zeileis@R-project.org
#     url: https://www.zeileis.org/
#   - name: Second  Author
#     affiliations:
#       - Plus Affiliation
# abstract: |
#   This short article illustrates how to write a manuscript for the
#   _Journal of Statistical Software_ (JSS) using its {{< latex >}} style files.
#   Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
#   it is recommended to keep the {{< latex >}} code as simple as possible,
#   i.e., avoid inclusion of packages/commands that are not necessary.
#   For outlining the typical structure of a JSS article some brief text snippets are employed that have been inspired by @ZeileisKleiberJackman2008, discussing count data regression in [R]{.proglang}. Editorial comments and instructions are marked by vertical bars.
abstract: |
  This article introduces the [simodels]{.pkg} package for spatial interaction modelling in R. The package provides functions for estimating spatial interaction models, including unconstrained gravity models and models that are production-, attraction-, or doubly-constrained.

keywords: [JSS, style guide, comma-separated, not capitalized, R]
keywords-formatted: [JSS, style guide, comma-separated, not capitalized, "[R]{.proglang}"]
-->

## Ideas

- Conceptual diagram showing decision points when developing SIMs, e.g.:
  - Unipartite or bipartite?
  - Unconstrained, production-constrained, attraction-constrained, or doubly-constrained? Or other?
  - Calculation and parameter optimisation or regression model?
- Open access dataset?

@@

### Basic implementation of SIMs {#sec-base}

@@

## The simodels package {#sec-simodels}

### OD data preparation {#sec-od}

### Interaction calculation {#sec-interaction}

### Interaction modelling {#sec-models}

### Optimising/parameter estimation {#sec-optimisation}

See https://github.com/acteng/netgen/blob/main/odgen.md#optimising-the-value-of-beta for ideas on this.

### Network distance measures {#sec-network}

@@

### Commuter flows between large areas in Leeds, UK

```{r}
od = simodels::si_od_census
zones = simodels::si_zones
nrow(od)
sum(od$all)
sum(zones$all)
```

### Commuter flows between small areas in Leeds, UK

@@

```{r}
#| echo: false
#| eval: false
od_observed_o = od_observed |> 
  group_by(OA11CD) |> 
  summarise(n = sum(n))
od_observed_d = od_observed |> 
  group_by(wz11cd) |> 
  summarise(n = sum(n))
zones_joined = left_join(
  zones |>
    select(OA11CD, n_o) |>
    sf::st_drop_geometry(),
  od_observed_o
)
head(zones_joined)
cor(zones_joined$n_o, zones_joined$n)

# For destinations:
destinations_joined = left_join(
  destinations |>
    select(wz11cd, n_d) |>
    sf::st_drop_geometry(),
  od_observed_d
)
cor(destinations_joined$n_d, destinations_joined$n)
```

```{r}
# TODO: test if the max_dist arg actually makes it faster for large datasets
system.time({
  od_modelled = si_to_od(zones, destinations)
})
od_modelled = si_to_od(zones, destinations)
system.time({
  od_modelled_max_5km = si_to_od(zones, destinations, max_dist = 3000)
})
od_modelled_max_5km = si_to_od(zones, destinations, max_dist = 3000)
dim(od_modelled)
names(od_modelled)
dim(od_modelled_max_5km)
```
